The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-04 paper/ms.tex
Small changes throughout

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia} and future missions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the observed values in $J_z$ for the thick-disk orbit
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account. Intuition for the shape
of each panel is given in the text. We also plot one half the 95\uth{}
percentile minus the 5\uth{} percentile of each distribution as a horizontal
arrow anchored on the true $J_z$ value. We call this $\Delta J_z$ and will use
it (along with the similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to
empirically describe the error distribution. We see that $\Delta J_z$ roughly
corresponds to the distance from the true $J_z$ value to one of the modes of
the distribution of observed $J_z$ values.}
\label{fig:Jz_hist}
\end{figure}

@@

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to
\sgra{} can be precisely measured \citep{2009ApJ...692.1075G,
2018AA...615L..15G}, with a recent measurement using near-infrared
interferometry of $8.178 \pm 0.035$~kpc \citep{2019arXiv190405721A}.

@@

\subsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of
stars near the Sun relative to the Galactic center (which is defined to have
zero velocity), and the velocity of the Sun relative to the LSR. The radial
($U_{\odot}$) and vertical ($W_{\odot}$) components are computed by taking the
mean motions of different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}.
The azimuthal component ($V_{\odot}$) is more difficult to measure, but can be
modelled using the asymmetric drift relation \citep{2008gady.book.....B}. The
values of the components of the LSR are usually taken from
\citet{2010MNRAS.403.1829S}. The value of the circular velocity is taken to be
$\sim 220\kms$ \citep[e.g.][]{2012ApJ...759..131B}.

@@

Considerable effort has been placed on each of these measurements, but
uncertainties remain, and detailed dynamical modelling across the disk ---
particularly for dynamically cold stars --- may have to take them into
account. For instance, the discrepancy of $\sim 5\,\pc$ between the midplane
defined by stars and by gas will induce a $\sim10\%$ error in $J_z$ for an
orbit with $z_{\text{max}}\sim100\,\pc$ (see Section~\ref{sec:ref_frame}).

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the Milky Way and Sagittarius are
each initialized to be in equilibrium in isolation \Gus{check!}.

@@

\item Dynamical modelling across large regions of the disk, over which the
midplane location varies by more than the limits discussed above, is
susceptible to this type of systematic error, since the assumption that our
local Galactic midplane is the global Galactic midplane is not true {\em a
priori}.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modeling across the large
regions of the disk accessible to {\em Gaia} and future missions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the observed values in $J_z$ for the thick-disk (top panel) and thin-disk (bottom panel) orbits
assuming a $z$ offset of $100\,\pc$. One can see that if the observed $z$
values have a bias (from e.g. an incorrectly computed midplane), then the
induced error distribution in $J_z$ is decidedly non-Gaussian. Therefore, any
sort of error propagation must take this into account. Intuition for the shape
of each panel is given in the text. We also plot one half the 95\uth{}
percentile minus the 5\uth{} percentile of each distribution as a horizontal
arrow anchored on the true $J_z$ value. We call this $\Delta J_z$ and will use
it (along with the similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to
empirically describe the error distribution. We see that $\Delta J_z$ roughly
corresponds to the distance from the true $J_z$ value to one of the modes of
the distribution of observed $J_z$ values.}
\label{fig:Jz_hist}
\end{figure}

@@

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to
\sgra{}, referred to as $R_0$, can be precisely measured \citep{2009ApJ...692.1075G,
2018AA...615L..15G}, with a recent measurement using near-infrared
interferometry of $8.178 \pm 0.035$~kpc \citep{2019arXiv190405721A}.

@@

\subsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of
stars near the Sun relative to the Galactic center (which is defined to have
zero velocity), and the velocity of the Sun relative to the LSR. The radial
($U_{\odot}$) and vertical ($W_{\odot}$) components are computed by taking the
mean motions of different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}.
The azimuthal component ($V_{\odot}$) is more difficult to measure, but can be
modeled using the asymmetric drift relation \citep{2008gady.book.....B}. The
values of the components of the LSR are usually taken from
\citet{2010MNRAS.403.1829S}. The value of the circular velocity is taken to be
$\sim 220\kms$ \citep[e.g.][]{2012ApJ...759..131B}.

@@

Considerable effort has been placed on each of these measurements, but
uncertainties remain, and detailed dynamical modeling across the disk ---
particularly for dynamically cold stars --- may have to take them into
account. For instance, the discrepancy of $\sim 5\,\pc$ between the midplane
defined by stars and by gas will induce a $\sim10\%$ error in $J_z$ for an
orbit with $z_{\text{max}}\sim100\,\pc$ (see Section~\ref{sec:ref_frame}).

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modeled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the Milky Way and Sagittarius are
each initialized to be in equilibrium in isolation \Gus{check!}.

@@

\item Dynamical modeling across large regions of the disk, over which the
midplane location varies by more than the limits discussed above, is
susceptible to this type of systematic error, since the assumption that our
local Galactic midplane is the global Galactic midplane is not true {\em a
priori}.
